---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(foist)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 11
)
```
# FOIst - Fast Output of Images

<!-- badges: start -->
![](https://img.shields.io/badge/Status-alpha-orange.svg)
![](https://img.shields.io/badge/Version-0.1.1-blue.svg)
<!-- badges: end -->


#### `foist` is a very fast way to output a matrix or array to a lossless image file.

**`foist` writes lossless grey image files ~5.5x faster than the `png` library.**

**`foist` writes lossless RGB image files ~7.5x faster than the `png` library**

**`foist` allocates 300-1000x *less* memory than png.**

* Only supports writing (lossless) [NETPBM](http://netpbm.sourceforge.net/)  PGM (grey) and PPM (RGB) files
* Only supports 8-bits-per-channel grey and RGB images.
* Uses [Rcpp](https://cran.r-project.org/package=Rcpp) to do 
    * **type conversion** from *double* to *unsigned byte*, and 
    * **data re-ordering** from R's column-major ordering to image output with row-major ordering
* By specifying `convert_to_row_major = FALSE` the image data-ordering is not converted
  from R's native data ordering - this makes image saving very fast as data manipulation
  operations are minimised and cache coherency is much improved.  However, the
  image will appear tranposed in the output.



## What's in the box

* `write_pgm()` - write a matrix to a grey image
* `write_ppm()` - write an array to an RGB image
* `write_pal_ppm()` write a matrix to an RGB image using a supplied palette
* `vir` 5 palettes from [viridis](https://cran.r-project.org/package=viridis) in the appropriate format

This package would not be possible without:

* [Rcpp](https://cran.r-project.org/package=Rcpp) - The easiest way to get fast C/C++ code into R
* [viridis](https://cran.r-project.org/package=viridis) - The wonderful palettes originally from [matplotlib](http://matplotlib.org)
* [NETPBM](http://netpbm.sourceforge.net) - A 30-year-old, rock-solid, uncompressed image format


## Caveats

Don't look at my C code unless you'd like a heart attack - or perhaps to lend a hand!

## Installation

You can install the development version from [GitHub](https://github.com/coolbutuseless/foist) with:

``` r
# install.packages("devtools")
devtools::install_github("coolbutuseless/foist")
```

## Setup data

* `dbl_mat` - A numeric matrix to output to a grey image. All values in range [0, 1]
* `dbl_arr` - A 3d numeric array to output to an RGB image. All values in range [0, 1]

```{r}
ncol    <- 256
nrow    <- 160
int_vec <- seq(nrow * ncol) %% 254L
int_mat <- matrix(int_vec, nrow = nrow, ncol = ncol, byrow = TRUE)
dbl_mat <- int_mat/255

# A non-boring RGB array/image
r       <- dbl_mat
g       <- matrix(rep(seq(0, 255, length.out = nrow)/255, each = ncol), nrow, ncol, byrow = TRUE)
b       <- dbl_mat[, rev(seq(ncol(dbl_mat)))  ]
dbl_arr <- array(c(r, g, b), dim = c(nrow, ncol, 3))
```


## Save a *2D matrix* as a grey image

A 2D numeric **matrix** can be saved as a grey image using `write_pgm()`.

The matrix values must be in the range [0, 1]. Use the `intensity_factor` argument to 
scale image values on-the-fly as they are written to file.

```{r example_pgm}
write_pgm(dbl_mat, "man/figures/col-0-n.pgm")
write_pgm(dbl_mat, "man/figures/col-0-f.pgm", flipy = TRUE)
write_pgm(dbl_mat, "man/figures/col-0-t.pgm", convert_to_row_major = FALSE)
```

```{r echo = FALSE}
system("convert man/figures/col-0-n.pgm -define png:exclude-chunks=date,time man/figures/col-convert-0-n.png")
system("convert man/figures/col-0-f.pgm -define png:exclude-chunks=date,time man/figures/col-convert-0-f.png")
system("convert man/figures/col-0-t.pgm -define png:exclude-chunks=date,time man/figures/col-convert-0-t.png")
```


<div>
  <img src = "man/figures/col-convert-0-n.png"  width = "30%" title = "no scaling">
  <img src = "man/figures/col-convert-0-f.png"  width = "30%" title = "flipy = TRUE"            style = "margin-left: 30px;">
  <img src = "man/figures/col-convert-0-t.png"  width = "19%" title = "intensity_factor = 0.5"  style = "margin-left: 30px;">
</div>


## Save a *3D array* as an RGB image 

An NxMx3 **array** can be saved as an RGB image. Each of the colours is represented
of one of the 3 planes of the array.

The array values must be in the range [0, 1]. Use the `intensity_factor` argument to 
scale image values on-the-fly as they are written to file.

```{r example_ppm}
write_ppm(dbl_arr, filename = "man/figures/col-1-n.ppm")
write_ppm(dbl_arr, filename = "man/figures/col-1-f.ppm", flipy = TRUE)
write_ppm(dbl_arr, filename = "man/figures/col-1-t.ppm", convert_to_row_major = FALSE)
```

```{r echo = FALSE}
system("convert man/figures/col-1-n.ppm -define png:exclude-chunks=date,time man/figures/col-convert-1-n.png")
system("convert man/figures/col-1-f.ppm -define png:exclude-chunks=date,time man/figures/col-convert-1-f.png")
system("convert man/figures/col-1-t.ppm -define png:exclude-chunks=date,time man/figures/col-convert-1-t.png")
```


<div>
  <img src = "man/figures/col-convert-1-n.png"  width = "30%" title = "convert_to_row_major = TRUE">
  <img src = "man/figures/col-convert-1-f.png"  width = "30%" title = "flipy = TRUE"                  style = "margin-left: 30px;">
  <img src = "man/figures/col-convert-1-t.png"  width = "19%" title = "convert_to_row_major = FALSE"  style = "margin-left: 30px;">
</div>


## Save a *matrix* to an RGB image using a palette lookup

Using `write_pal_ppm()`, `foist` can write a grey image as an RGB image by using 
each grey pixel value to lookup an RGB colour in a given palette.

A palette must be an integer matrix with dimensions 256 x 3 and values in the
range [0, 255].

The matrix values must be in the range [0, 1]. Use the `intensity_factor` argument to 
scale image values on-the-fly as they are written to file.

`foist` includes the 5 palettes from [viridis](https://cran.r-project.org/package=viridis) as `vir$magma` etc.

```{r example_ppm_from_matrix}
foist::write_pgm    (dbl_mat,                           "man/figures/col-0.pgm")
foist::write_pal_ppm(dbl_mat, pal = foist::vir$magma  , "man/figures/col-3.ppm")
foist::write_pal_ppm(dbl_mat, pal = foist::vir$inferno, "man/figures/col-4.ppm")
foist::write_pal_ppm(dbl_mat, pal = foist::vir$plasma , "man/figures/col-5.ppm")
foist::write_pal_ppm(dbl_mat, pal = foist::vir$viridis, "man/figures/col-6.ppm")
foist::write_pal_ppm(dbl_mat, pal = foist::vir$cividis, "man/figures/col-7.ppm")
```

```{r echo = FALSE}
system("convert man/figures/col-3.ppm -define png:exclude-chunks=date,time man/figures/col-convert-3.png")
system("convert man/figures/col-4.ppm -define png:exclude-chunks=date,time man/figures/col-convert-4.png")
system("convert man/figures/col-5.ppm -define png:exclude-chunks=date,time man/figures/col-convert-5.png")
system("convert man/figures/col-6.ppm -define png:exclude-chunks=date,time man/figures/col-convert-6.png")
system("convert man/figures/col-7.ppm -define png:exclude-chunks=date,time man/figures/col-convert-7.png")
```

<div>
  <img src = "man/figures/col-convert-0-n.png" width = "30%" title = "grey">
  <img src = "man/figures/col-convert-3.png"   width = "30%" title = "magma">
  <img src = "man/figures/col-convert-4.png"   width = "30%" title = "inferno">
  <img src = "man/figures/col-convert-5.png"   width = "30%" title = "plasma">
  <img src = "man/figures/col-convert-6.png"   width = "30%" title = "viridis">
  <img src = "man/figures/col-convert-7.png"   width = "30%" title = "cividis">
</div>







```{r echo = FALSE}
library(dplyr)
library(ggplot2) 

ncol    <- 1024
nrow    <- 640
int_vec <- rep.int(seq(ncol) - 1, nrow) %% 256L
int_mat <- matrix(int_vec, nrow = nrow, ncol = ncol, byrow = TRUE)
dbl_mat <- int_mat/255

r       <- dbl_mat
g       <- matrix(rep(seq(0, 255, length.out = nrow)/255, each = ncol), nrow, ncol, byrow = TRUE)
b       <- dbl_mat[, rev(seq(ncol(dbl_mat)))  ]

dbl_arr <- array(c(r, g, b), dim = c(nrow, ncol, 3))
```




## Benchmark: Saving a matrix as a grey image


The following benchmark compares:

* `foist::write_pgm()` in both row-major and column-major ordering
    * by **not** converting to row-major ordering the data is written in  the same order it 
      is stored in R.  By minimizing this data manipulation some significant 
      speedups are achieved.
* `png::writePNG()`
* `jpeg::writeJPEG()` - I'm not actually interested in lossy output, 
  but it is interesting to note just how fast the jpeg library is.


```{r}
tmp <- tempfile()

res <- bench::mark(
  `foist::write_pgm()`                    = foist::write_pgm(dbl_mat, tmp),
  `foist::write_pgm(column-major)`        = foist::write_pgm(dbl_mat, tmp, convert_to_row_major = FALSE),
  `foist::write_pgm(column-major, flipy)` = foist::write_pgm(dbl_mat, tmp, convert_to_row_major = FALSE, flipy = TRUE),
  `png::writePNG()`                       = png::writePNG   (dbl_mat, tmp),
  `jpeg::writeJPEG`                       = jpeg::writeJPEG (dbl_mat, tmp),
  min_time = 2, check = FALSE
)
```


```{r benchmark_grey, echo = FALSE}
res %>%
  select(expression, min, mean, median, `itr/sec`, mem_alloc) %>%
  mutate(`itr/sec` = round(`itr/sec`)) %>%
  knitr::kable(caption = "Benchmark results")

plot(res) + 
  theme_bw(15) + 
  theme(legend.position = 'bottom') + 
  labs(title = 'Saving a 1024x640 grey image', y = '') 
```




## Benchmark: Saving an array as an RGB image

The following benchmark compares:

* `foist::write_ppm()` in both row-major and column-major ordering
    * by **not** converting to row-major ordering the data is written in  the same order it 
      is stored in R.  By minimizing this data manipulation some significant 
      speedups are achieved.
* `png::writePNG()`
* `jpeg::writeJPEG()` - I'm not actually interested in lossy output, 
  but it is interesting to note just how fast the jpeg library is.




```{r}
tmp <- tempfile()

res <- bench::mark(
  `foist::write_ppm()`                    = foist::write_ppm(dbl_arr, tmp),
  `foist::write_ppm(column-major)`        = foist::write_ppm(dbl_arr, tmp, convert_to_row_major = FALSE),
  `foist::write_ppm(column-major, flipy)` = foist::write_ppm(dbl_arr, tmp, convert_to_row_major = FALSE, flipy = TRUE),
  `png::writePNG()`                       = png::writePNG   (dbl_arr, tmp),
  `jpeg::writeJPEG`                       = jpeg::writeJPEG (dbl_arr, tmp),
  min_time = 2, check = FALSE
)
```


```{r benchmark_rgb, echo = FALSE}
res %>%
  select(expression, min, mean, median, `itr/sec`, mem_alloc) %>%
  mutate(`itr/sec` = round(`itr/sec`)) %>%
  knitr::kable(caption = "Benchmark results")

plot(res) + 
  theme_bw(15) + 
  theme(legend.position = 'bottom') + 
  labs(title = 'Saving a 1024x640 RGB image', y = '')
```

















