---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(foist)
})

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.width = 11
)
```
# FOIst - Fast Output of Images

<!-- badges: start -->
![](https://img.shields.io/badge/Status-alpha-orange.svg)
![](https://img.shields.io/badge/Version-0.1.3-blue.svg)
![](https://img.shields.io/badge/Output-PNG-green.svg)
![](https://img.shields.io/badge/Output-PGM-green.svg)
![](https://img.shields.io/badge/Output-PPM-green.svg)
<!-- badges: end -->


#### `foist` is a very fast way to output a matrix or array to a lossless image file.

<br/>

**`foist` writes lossless grey image files ~5.5x faster than the `png` library.**

**`foist` writes lossless RGB image files ~7.5x faster than the `png` library**

**`foist` allocates 300-1000x *less* memory than png.**

<br/>

* Only supports writing (lossless) images:
    * [NETPBM](http://netpbm.sourceforge.net/)  PGM (grey) and PPM (RGB) files
    * [PNG](https://en.wikipedia.org/wiki/Portable_Network_Graphics) files - Grey, RGB or with a fixed 256 colour palette.
* Only supports 8-bits-per-channel grey and RGB images.
* Uses [Rcpp](https://cran.r-project.org/package=Rcpp) to do 
    * type conversion from *double* to *unsigned byte*, and 
    * data re-ordering from R's column-major ordering to image output with row-major ordering
* By specifying `convert_to_row_major = FALSE` the image data-ordering is not converted
  from R's native data ordering - this makes image saving very fast as data manipulation
  operations are minimised and cache coherency is much improved.  However, the
  image will appear tranposed in the output.



## What's in the box

* `write_pnm()` and `write_png()` which can
    - write an array to an RGB image
    - write a matrix to a grey image
    - write a matrix to a RGB image with a supplied palette
* `vir` 5 palettes from [viridis](https://cran.r-project.org/package=viridis) in the appropriate format

This package would not be possible without:

* [Rcpp](https://cran.r-project.org/package=Rcpp) - The easiest way to get fast C/C++ code into R
* [viridis](https://cran.r-project.org/package=viridis) - The wonderful palettes originally from [matplotlib](http://matplotlib.org)
* [NETPBM](http://netpbm.sourceforge.net) - A 30-year-old, rock-solid, uncompressed image format
* [PNG](https://www.w3.org/TR/PNG/)


## Caveats

Don't look at my C/C++ code unless you'd like a heart attack.

## Technical Notes

* `foist` contains a bespoke PNG encoder I wrote in C++
    * Encoder uses Mark Adler's `adler32.c` code from [zlib](https://www.zlib.net/) Copyright (C) 1995-2011, 2016 Mark Adler
    * A SIMD version of `adler32()` is included but not enabled by default. See `#define ADLER_SIMD` in `write-png.cpp`
    * `crc32` implementation is a very fast slice-by-16 implementation by [Stephan Brumme](https://create.stephan-brumme.com/crc32/).
      This is noticeably much faster than the slice-by-4 crc32 that comes with the standard [zlib library](https://www.zlib.net/) 
      or in the R package [digest](https://cran.r-project.org/package=digest)
* When converting the numeric values to unsigned bytes for image output, they 
  are truncated to integer rather than rounded

## Installation

You can install the development version from [GitHub](https://github.com/coolbutuseless/foist) with:

``` r
# install.packages("devtools")
devtools::install_github("coolbutuseless/foist")
```

## Setup data

* `dbl_mat` - A numeric matrix to output to a grey image. All values in range [0, 1]
* `dbl_arr` - A 3d numeric array to output to an RGB image. All values in range [0, 1]

```{r}
ncol    <- 256
nrow    <- 160
int_vec <- seq(nrow * ncol) %% 254L
int_mat <- matrix(int_vec, nrow = nrow, ncol = ncol, byrow = TRUE)
dbl_mat <- int_mat/255

# A non-boring RGB array/image
r       <- dbl_mat
g       <- matrix(rep(seq(0, 255, length.out = nrow)/255, each = ncol), nrow, ncol, byrow = TRUE)
b       <- dbl_mat[, rev(seq(ncol(dbl_mat)))  ]
dbl_arr <- array(c(r, g, b), dim = c(nrow, ncol, 3))
```


## Save a *2D matrix* as a grey image

A 2D numeric **matrix** can be saved as a grey image..

The matrix values must be in the range [0, 1]. Use the `intensity_factor` argument to 
scale image values on-the-fly as they are written to file.

```{r example_pgm}
# NETPBM PGM
write_pnm(dbl_mat, "man/figures/col-0-n.pgm")
write_pnm(dbl_mat, "man/figures/col-0-f.pgm", flipy = TRUE)
write_pnm(dbl_mat, "man/figures/col-0-t.pgm", convert_to_row_major = FALSE)

# PNG
write_png(dbl_mat, "man/figures/col-0-n.png")
write_png(dbl_mat, "man/figures/col-0-f.png", flipy = TRUE)
write_png(dbl_mat, "man/figures/col-0-t.png", convert_to_row_major = FALSE)

```

```{r echo = FALSE}
system("convert man/figures/col-0-n.pgm -define png:exclude-chunks=date,time man/figures/col-convert-0-n.png")
system("convert man/figures/col-0-f.pgm -define png:exclude-chunks=date,time man/figures/col-convert-0-f.png")
system("convert man/figures/col-0-t.pgm -define png:exclude-chunks=date,time man/figures/col-convert-0-t.png")
```

<div>
  <img src = "man/figures/col-convert-0-n.png"  width = "30%" title = "no scaling">
  <img src = "man/figures/col-convert-0-f.png"  width = "30%" title = "flipy = TRUE"            style = "margin-left: 30px;">
  <img src = "man/figures/col-convert-0-t.png"  width = "19%" title = "intensity_factor = 0.5"  style = "margin-left: 30px;">
</div>



## Save a *3D array* as an RGB image 

An NxMx3 **array** can be saved as an RGB image. Each of the colours is represented
of one of the 3 planes of the array.

The array values must be in the range [0, 1]. Use the `intensity_factor` argument to 
scale image values on-the-fly as they are written to file.

```{r example_ppm}
# NETPBM PPM format
write_pnm(dbl_arr, filename = "man/figures/col-1-n.ppm")
write_pnm(dbl_arr, filename = "man/figures/col-1-f.ppm", flipy = TRUE)
write_pnm(dbl_arr, filename = "man/figures/col-1-t.ppm", convert_to_row_major = FALSE)

# PNG
write_pnm(dbl_arr, filename = "man/figures/col-1-n.png")
write_pnm(dbl_arr, filename = "man/figures/col-1-f.png", flipy = TRUE)
write_pnm(dbl_arr, filename = "man/figures/col-1-t.png", convert_to_row_major = FALSE)
```

```{r echo = FALSE}
system("convert man/figures/col-1-n.ppm -define png:exclude-chunks=date,time man/figures/col-convert-1-n.png")
system("convert man/figures/col-1-f.ppm -define png:exclude-chunks=date,time man/figures/col-convert-1-f.png")
system("convert man/figures/col-1-t.ppm -define png:exclude-chunks=date,time man/figures/col-convert-1-t.png")
```

<div>
  <img src = "man/figures/col-convert-1-n.png"  width = "30%" title = "convert_to_row_major = TRUE">
  <img src = "man/figures/col-convert-1-f.png"  width = "30%" title = "flipy = TRUE"                  style = "margin-left: 30px;">
  <img src = "man/figures/col-convert-1-t.png"  width = "19%" title = "convert_to_row_major = FALSE"  style = "margin-left: 30px;">
</div>



## Save a *matrix* to an RGB image using a palette lookup

`foist` can write a grey image as an RGB image by using 
each grey pixel value to lookup an RGB colour in a given palette.

Using `write_png()` to write a paletted image will be faster than `write_pnm()`
because the PNG format directly supports palettes.

A palette must be an integer matrix with dimensions 256 x 3 and values in the
range [0, 255].

`foist` includes the 5 palettes from [viridis](https://cran.r-project.org/package=viridis) as `vir$magma` etc.

```{r example_ppm_from_matrix}
# NETPBM format
foist::write_pnm(dbl_mat,                           "man/figures/col-0.pgm")
foist::write_pnm(dbl_mat, pal = foist::vir$magma  , "man/figures/col-3.ppm")
foist::write_pnm(dbl_mat, pal = foist::vir$inferno, "man/figures/col-4.ppm")
foist::write_pnm(dbl_mat, pal = foist::vir$plasma , "man/figures/col-5.ppm")
foist::write_pnm(dbl_mat, pal = foist::vir$viridis, "man/figures/col-6.ppm")
foist::write_pnm(dbl_mat, pal = foist::vir$cividis, "man/figures/col-7.ppm")

# PNG format
foist::write_png(dbl_mat,                           "man/figures/col-0.png")
foist::write_png(dbl_mat, pal = foist::vir$magma  , "man/figures/col-3.png")
foist::write_png(dbl_mat, pal = foist::vir$inferno, "man/figures/col-4.png")
foist::write_png(dbl_mat, pal = foist::vir$plasma , "man/figures/col-5.png")
foist::write_png(dbl_mat, pal = foist::vir$viridis, "man/figures/col-6.png")
foist::write_png(dbl_mat, pal = foist::vir$cividis, "man/figures/col-7.png")
```

```{r echo = FALSE}
system("convert man/figures/col-3.ppm -define png:exclude-chunks=date,time man/figures/col-convert-3.png")
system("convert man/figures/col-4.ppm -define png:exclude-chunks=date,time man/figures/col-convert-4.png")
system("convert man/figures/col-5.ppm -define png:exclude-chunks=date,time man/figures/col-convert-5.png")
system("convert man/figures/col-6.ppm -define png:exclude-chunks=date,time man/figures/col-convert-6.png")
system("convert man/figures/col-7.ppm -define png:exclude-chunks=date,time man/figures/col-convert-7.png")
```

<div>
  <img src = "man/figures/col-convert-0-n.png" width = "30%" title = "grey">
  <img src = "man/figures/col-convert-3.png"   width = "30%" title = "magma">
  <img src = "man/figures/col-convert-4.png"   width = "30%" title = "inferno">
  <img src = "man/figures/col-convert-5.png"   width = "30%" title = "plasma">
  <img src = "man/figures/col-convert-6.png"   width = "30%" title = "viridis">
  <img src = "man/figures/col-convert-7.png"   width = "30%" title = "cividis">
</div>







```{r echo = FALSE}
library(dplyr)
library(ggplot2) 

ncol    <- 1024
nrow    <- 640
int_vec <- rep.int(seq(ncol) - 1, nrow) %% 256L
int_mat <- matrix(int_vec, nrow = nrow, ncol = ncol, byrow = TRUE)
dbl_mat <- int_mat/255

r       <- dbl_mat
g       <- matrix(rep(seq(0, 255, length.out = nrow)/255, each = ncol), nrow, ncol, byrow = TRUE)
b       <- dbl_mat[, rev(seq(ncol(dbl_mat)))  ]

dbl_arr <- array(c(r, g, b), dim = c(nrow, ncol, 3))
```




## Benchmark: Saving a matrix as a grey image


The following benchmark compares the time to output of a grey image using:

* `foist::write_pnm()` in both row-major and column-major ordering
    * by **not** converting to row-major ordering the data is written in  the same order it 
      is stored in R.  By minimizing this data manipulation some significant 
      speedups are achieved.
* `png::writePNG()`


```{r}
tmp <- tempfile()

res <- bench::mark(
  `foist::write_pnm()`                    = foist::write_pnm(dbl_mat, tmp),
  `foist::write_pnm(column-major)`        = foist::write_pnm(dbl_mat, tmp, convert_to_row_major = FALSE),
  
  `foist::write_png()`                    = foist::write_png(dbl_mat, tmp),
  `foist::write_png(column-major)`        = foist::write_png(dbl_mat, tmp, convert_to_row_major = FALSE),
  
  `png::writePNG()`                       = png::writePNG   (dbl_mat, tmp),
  min_time = 2, check = FALSE
)
```


```{r benchmark_grey, echo = FALSE}
res %>%
  select(expression, min, mean, median, `itr/sec`, mem_alloc) %>%
  mutate(`itr/sec` = round(`itr/sec`)) %>%
  knitr::kable(caption = "Benchmark results")

plot(res) + 
  theme_bw(15) + 
  theme(legend.position = 'bottom') + 
  labs(title = 'Saving a 1024x640 grey image', y = '') 
```




## Benchmark: Saving an array as an RGB image

The following benchmark compares the time to output a colour image using:

* `foist::write_pnm()` in both row-major and column-major ordering
    * by **not** converting to row-major ordering the data is written in  the same order it 
      is stored in R.  By minimizing this data manipulation some significant 
      speedups are achieved.
* `png::writePNG()`




```{r}
tmp <- tempfile()

res <- bench::mark(
  `foist::write_pnm()`                    = foist::write_pnm(dbl_arr, tmp),
  `foist::write_pnm(column-major)`        = foist::write_pnm(dbl_arr, tmp, convert_to_row_major = FALSE),
  
  `foist::write_png()`                    = foist::write_png(dbl_arr, tmp),
  `foist::write_png(column-major)`        = foist::write_png(dbl_arr, tmp, convert_to_row_major = FALSE),
  
  `png::writePNG()`                       = png::writePNG   (dbl_arr, tmp),
  min_time = 2, check = FALSE
)
```


```{r benchmark_rgb, echo = FALSE}
res %>%
  select(expression, min, mean, median, `itr/sec`, mem_alloc) %>%
  mutate(`itr/sec` = round(`itr/sec`)) %>%
  knitr::kable(caption = "Benchmark results")

plot(res) + 
  theme_bw(15) + 
  theme(legend.position = 'bottom') + 
  labs(title = 'Saving a 1024x640 RGB image', y = '')
```

















